\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-06 analysis/paper/paper.Rmd
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
library(splines)
library(tibble)
library(glmnet)
library(kableExtra)
library(pracma)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
```

@@

Survival analysis has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression]. This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model. This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates. To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion]. This in turn often leads to stepwise estimates of the survival function.

From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of having a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group. Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can then be used to estimate the cumulative incidence function (CIF). From this perspective, the CIF and the survival function estimates are now smooth functions of time.

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their approach provides users that are familiar with generalized linear models with a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of partial likelihood approaches.

In this article, we present an \proglang{R} package that combines the ideas of Hanley & Miettinen into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parameteric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures. Finally, we end the article with a discussion of the results and of future directions.

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest sampling the base series *uniformly* from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

<!-- The `casebase` package fits the family of hazard functions of the form -->

<!-- $$ h(t;X) = \exp[g(t;X)] $$ -->
<!-- where $t$ denotes time and $X$, the individual's covariate profile.  -->

Let $g(t; X)$ be the linear predictor such that $\log(h(t;X)) = g(t; X)$. Different functions of $t$ lead to different parametric hazard models. The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$:

$$ \log(h(t; X)) = \beta_0 + \beta_1 X.$$

The instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \textit{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

$$ \log(h(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$

Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time [@kalbfleisch2011statistical]:

$$ \log(h(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$

@@

where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

In the next case study, we show how case-base sampling can  be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.

```{r bmtcrr-data, eval = eval_cs2, message = FALSE}
library(casebase)
data(bmtcrr)
```

@@

First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red dots for the event of interest (panel A) and blue dots for competing events (panel B). In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

@@

Our main objective is to compute the absolute risk of relapse for a given set of covariates. We start by fitting a smooth hazard to the data using a linear term for time:

@@

table_cb %>% 
  dplyr::inner_join(table_cox, by = "Covariates") %>%
  dplyr::mutate(Covariates = dplyr::case_when(
    Covariates == "SexM" ~ "Sex",
    Covariates == "DAML" ~ "Disease",
    Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
    Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
    Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
    Covariates == "SourcePB" ~ "Source",
    TRUE ~ Covariates
    )) %>% 
  knitr::kable(format = "latex", booktabs = TRUE,
               col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
               caption = "Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.") %>% 
  kable_styling() %>% 
  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
```

@@

We will compare our estimates to that obtained from a corresponding Fine-Gray model [-@fine1999proportional]. The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e. the function $f_k(t)$ such that
$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
where $CI_k(t)$ is the cause-specific cumulative indicence. The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard. For the computation, we will use the \pkg{timereg} package [@timereg]:

@@

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using predictive mean matching and using the \pkg{mice} package in \proglang{R} [@mice]. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 individuals. \textcolor{red}{How come we lose individuals during imputation?} Of these, 67.5% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the website of the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}. \textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death")
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death and cannot be used as a predictor. Then we select one observation at random. We will estimate the cumulative incidence curve for this individual. The other observations will be used to fit the hazard function.

```{r supportSetUp, echo = TRUE, eval = eval_cs3}
support <- dplyr::select(support, -hospdead)

# Randomly select one observation
indiv <- sample(1:nrow(support), size = 1)
```

Our main objective is to estimate the cumulative incidence of death for a given covariate profile. Given the number of covariates available, we will do this using regularized estimation. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will use the function \code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}. Specifically, we input a matrix \code{y}, which contains the time and event variables, and a matrix \code{x}, which contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables.

```{r support-design, eval = eval_cs3}
# Create design matrix without the time variable
support_matrix <- model.matrix(death ~ . - d.time, 
                               data = support)
# Remove intercept term
support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]

# Create test and train dataset
support_test <- support_matrix[indiv,,drop = FALSE]

support_train <- support_matrix[-indiv,]
y <- as.matrix(support[-indiv, c("death", "d.time")])
```

We are now ready to estimate the hazard function. We use lasso penalization; this is achieved by setting \code{alpha} to 1. Also, note that the argument \code{formula_time} can be used to specify the function form for time; in our example, we model time linearly.

```{r CBhazard, echo = TRUE, eval = eval_cs3}
support_cb <- fitSmoothHazard.fit(support_train, y, 
                                  formula_time = ~ d.time, 
                                  family = "glmnet", 
                                  event = "death", time = "d.time", 
                                  alpha = 1, ratio = 100)
```

Then, we take our model object and use the \code{absoluteRisk} function to compute our estimate of the cumulative incidence function. Since \code{glmnet} fitted the penalized logistic regression for a range of penalty terms, we need to specify which one we want to use for our prediction. We select \code{s = "lambda.1se"}, as recommended.

```{r CBabsolute, echo = TRUE, eval = eval_cs3}
time_points <- seq(1, max(support$d.time), by = 0.5)
support_risk <- absoluteRisk(support_cb,
                             time = time_points,
                             newdata = support_test,
                             s = "lambda.1se", method = "numerical")
```

We will compare this curve to a regularized version of Cox regression, and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we can fit the Cox regression model using regularized estimation.

```{r coxHaz, echo = TRUE, eval = eval_cs3}
# Create survival object for glmnet
surv_glmnet <- survival::Surv(time = y[, "d.time"], 
                              event = y[, "death"])

support_cox <- glmnet::cv.glmnet(x = support_train, 
                                 y = surv_glmnet,
                                 family = "cox", alpha = 1)
```

However, \pkg{glmnet} does not provide any functionality for estimating the baseline hazard, and the functionalities in the \pkg{survival} package do not work on the output from \code{cv.glmnet}. To circumvent this issue and get an estimate of the baseline hazard, we followed the following steps. First, we fitted a non-regularized Cox model using only the variables selected by \pkg{glmnet}; the baseline hazard will be estimated using this object. Then, we replaced the estimates of the regression coefficients by the estimates from \pkg{glmnet}. Finally, this modified \code{coxph} object is fed to the \code{survival::survfit} function to obtain the estimated survival function.

```{r coxAbsolute, echo = TRUE, eval = eval_cs3}
# Identifying selected covariates
nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
                                  s = "lambda.1se")
nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
# Creating a new dataset that only contains the covariates selected by glmnet
support_cox_data <- as.data.frame(cbind(y,
                                        support_train[, nonzero_covariates_cox$X1]))
# Fitting a cox model
support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
                          data = support_cox_data)
# Replace the estimate with those obtained by glmnet
support_cox$coefficients <- nonzero_coef_cox@x

# Fitting CI curve for cox+glmnet
support_cox_risk <- survival::survfit(support_cox,
                                      newdata = as.data.frame(support_test),
                                      time = time_points,
                                      type = "breslow")
```

We used the \pkg{survival} package to calculate the Kaplan-Meier estimate of the survival function.

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}
km <- survival::Surv(time = support$d.time, 
                     event = support$death)
abKm <- survival::survfit(km ~ 1,
                          type = 'kaplan-meier')
```

Now that our three cumulative incidence functions have been calculated, we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and \pkg{casebase} have a lower estimated cumulative incidence by the end of the study, when compared to Kaplan-Meier. \textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue."}
plot(support_risk, type='l', col = "black", lwd = 2, 
     main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
     xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
lines(support_cox_risk, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
legend("bottom", 
       legend = c("casebase", "Cox+glmnet", "KM curve"), 
       col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
```

@@

In the previous case studies, we only considered covariates that were fixed at baseline. In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-varying covariates. As an example that already appeared in the literature, case-base sampling was to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case]. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

Recall the setting of Stanford Heart Transplant study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searched for a suitable donor for the patient, which could take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.

@@

The non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, we can turn these estimates of the hazard function into estimates of the cumulative incidence functions.

@@

In this article, we presented the \proglang{R} package \pkg{casebase}, which provides functions for fitting smooth parametric hazards and estimating cumulative incidence functions using case-base sampling. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through four case studies.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(tidyverse)
library(magrittr)
library(survival)
library(casebase)
#devtools::load_all('../../../casebase')#testing
library(splines)
library(tibble)
library(glmnet)
library(data.table)
library(kableExtra)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- FALSE
eval_theory <- FALSE
eval_implementation <- FALSE
eval_cs1 <- FALSE
eval_cs2 <- FALSE
eval_cs3 <- TRUE
eval_cs4 <- FALSE
```

@@

The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to predict the risk (or cumulative incidence (CI)) of an event, for a particular patient. The following points should be noted:

  1. hazard ratios
  2. If, however, the absolute risks are of interest, they have to be recovered using the semi-parametric Breslow estimator
  3. Alternative approaches for fitting flexible hazard models for estimating absolute risks, not requiring this two-step approach? Yes! [@hanley2009fitting]

[@hanley2009fitting] propose a fully parametric hazard model that can be fit via logistic regression. From the fitted hazard function, cumulative incidence and, thus, risk functions of time, treatment and profile can be easily derived.

@@

In Hanley & Miettinen [-@hanley2009fitting], the authors suggest performing case-base sampling *uniformly*, i.e. to sample the base series uniformly from the study base. In terms of Poisson processes, this sampling strategy corresponds essentially to a time-homogeneous Poisson process with intensity equal to $b/B$, where $b$ is the number of sampled observations in the base series, and $B$ is the total population-time for the study base. More complex examples are also available; see for example Saarela & Arjas [-@saarela2015non], where the probabilities of the sampling mechanism  are proportional to the cardiovascular disease event rate given by the Framingham score.

The `casebase` package fits the family of hazard functions of the form

$$ h(t;X) = \exp[g(t;X)] $$
where $t$ denotes time and $X$, the individual's covariate profile. Different functions of $t$ lead to different parametric hazard models. The simplest of these models is the one-parameter exponential distribution which is obtained by taking the hazard function to be constant over the range of $t$.

$$ h(t;X) = \exp(\beta_0 + \beta_1 X) $$

The instantaneous failure rate is independent of $t$, so that the conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the *memoryless property* (Kalbfleisch and Prentice, 2002).

@@

$$ h(t;X)  = \exp(\beta_0 + \beta_1 t + \beta_2 X) $$

Use of $\log(t)$ yields the Weibull hazard which allows for a power dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

$$ h(t;X)  = \exp(\beta_0 + \beta_1 \log(t) + \beta_2 X) $$

@@

where $N_j(t)$ is the counting process associated with the event of type $j$ and $h_j(t)$ is the corresponding hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1\end{align*}}

@@

In the next case study, we show how case-base sampling can also be used in the context of a competing risk analysis. For illustrative purposes, we will use the same data that was used in Scrucca *et al* [-@scrucca2010regression]. The data was downloaded from the first author's website, and it is also available as part of the \pkg{casebase} package.

```{r bmtcrr-data, eval = eval_cs2, message = FALSE}

data(bmtcrr)
```

@@

First, we can look at a population-time plot to visualize the incidence density of both relapse and the competing events. In Figure \ref{fig:compPop}, failure times are highlighted on the plot using red points for the event of interest (panel A) and blue dots for competing events (panel B). In both panels, we see evidence of a non-constant hazard function: the density of points is larger at the beginning of follow-up than at the end.

@@

Our main objective is to compute the absolute risk of relapse for a given set of covariates. First, we fit a smooth hazard to the data using a linear term for time:

@@

table_cb %>% 
  dplyr::inner_join(table_cox, by = "Covariates") %>%
  dplyr::mutate(Covariates = dplyr::case_when(
    Covariates == "SexM" ~ "Sex",
    Covariates == "DAML" ~ "Disease",
    Covariates == "PhaseCR2" ~ "Phase (CR2 vs. CR1)",
    Covariates == "PhaseCR3" ~ "Phase (CR3 vs. CR1)",
    Covariates == "PhaseRelapse" ~ "Phase (Relapse vs. CR1)",
    Covariates == "SourcePB" ~ "Source",
    TRUE ~ Covariates
    )) %>% 
  knitr::kable(format = "latex", booktabs = TRUE,
               col.names = c("Covariates", "HR", "95% CI", "HR", "95% CI"),
               caption = "ADD CAPTION") %>% 
  kable_styling() %>% 
  add_header_above(c(" " = 1, "Case-Base" = 2, "Cox" = 2))
```

@@

We will compare our estimates to that obtained from a corresponding Fine-Gray model [@fine1999proportional]. The Fine-Gray model is a semiparametric model for the cause-specific *subdistribution*, i.e. the function $f_k(t)$ such that
$$CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),$$
where $CI_k(t)$ is the cause-specific cumulative indicence. The Fine-Gray model allows to directly assess the effect of a covariate on the subdistribution, as opposed to the hazard. For the computation, we will use the \pkg{timereg} package [@timereg].

@@

For the third case study, we will introduce regularization into the estimation of the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 1000 individuals and 35 variables. charges, Totcst, totmcst and sfdm2 were removed as they are response variables we are not interested in for the sake of this analysis. After imputation, we have 690 individuals. We lose individuals due to coliniarity between variables, preventing imputation. If a pair of covariates are deemed colinear to a certain degree, they are excluded from any imputations. Here, only edu  income  avtisst*  race  ph*  glucose*  adlp* were imputed. Of these 690 individuals, 66.09% experienced the event of interest. <!--The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.--> A description of each variable can be found in Table \ref{tab:support1} and a breakdown of each categorical variable in Table \ref{tab:support2}. The original data was obtained from the Department of Biostatistics at Vanderbilt University. The imputed dataset is available as part of the \pkg{casebase} package.

@@

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.

```{r support-poptime, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death. All covariates will be used to estimate the hazard function. The hazard function will be used to estimate the cumulative incidence of each individual in the study. Then, the average cumulative incidience curve for all observations will be used for our comparison.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#remove hospital dead
completeData<-completeData[,-c(4)]
ratio=100  #universal ratio for eval_cs3
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])
```

Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data using a weibull distribution. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into indicator variables. Here, we used lasso penalization by setting alpha to 1.

```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}

cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                                       formula_time = ~log(d.time),alpha=1,ratio=ratio)
```
Then, we take our model and use the `absoluteRisk` function to integrate and retrieve our cumulative incidence curves.

```{r CBabsolute, echo = FALSE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseCI=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 20),newdata = x , s="lambda.1se",method=c("numerical"))

cbCIAve=data.frame(casebaseCI[,1],rowMeans(casebaseCI[,-c(1)]))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsolute, echo = FALSE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))
#hazard for cox using glmnet
coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
#plot(coxglmFit)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
#The coefficients of this object will be replaced with the estimates from coxglmFit.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxFit$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
newDataCox=newData[,nonzero_covariate_cox$X1-1]
abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(newDataCox),time = seq(0,max(completeData$d.time), 1),type="breslow")
abCF<-abCoxFit
abCF$surv<-rowMeans(abCF$surv)
```
We used the survival package to calculate cox regression's cumulative incidience without penalization and the kaplan-meier curve for incidience.

```{r unpenalizedCox, echo = FALSE, eval = eval_cs3}
Cox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,-c(1)]))
#coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
#abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
#abCFup<-abcoxFitup
#abCFup$surv<-rowMeans(abCFup$surv)
#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
       #                                ratio=rat,data=fullDataCox)
#summary(cb.ModelNorm)#need to figure out how to deal with NAs
#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))
```

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}

coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)

#creating a surv object to be used to fit an unadjusted absolute risk curve.
# (Kaplan-Meier risk curve)
km <- survival::Surv(time = completeData$d.time, event = completeData$death)
abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
```

Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
abCFup<-abcoxFitup
abCFup$surv<-rowMeans(abCFup$surv)
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(cbCIAve[,1],cbCIAve[,2],type='l',col="black",lwd=3,main="Cumulative Incidence comparison",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCF,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
lines(abCFup ,col="purple",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright", 
       legend = c( "Penalized Weibull with casebase","Coxnet","Cox","Kaplan-Meier"), 
       col = c("black","red","Blue","purple"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

```{r covariatesRegularizedLollipopPlot,eval=FALSE}

#get covariates excluding d.time
estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimates=rbind(estimatescb, estimatescox)
estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
        geom_point(position=ggstance::position_dodgev(height=0.2))

```

@@

Although the previous case studies provide a broad overview of the capabilities of \pkg{casebase}, they all have two properties in common:
\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case]. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can also be used for survival analysis problems that do not share the properties above. To this end, we will use the Stanford Heart Transplant Data [@clark1971cardiac,@crowley1977covariance]; we will use the version available in the \pkg{survival} package.

Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.

@@

The non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, we can turn these estimate of the hazard function into estimates of the cumulative incidence functions.

@@

In this article, we presented the \proglang{R} package \pkg{casebase} that provides functions to fit smooth parametric hazards and estimate cumulative incidence functions using case-base sampling. We outlined the theoretical underpinnings of the approach, we provided details about our implementation, and we illustrated the merits of the approach and the package through three case studies.

<!-- In the following table we provide a comparison between the Cox model and case-base sampling: -->

```{r, echo=FALSE, eval = FALSE, results='asis'}
knitr::kable(data.frame(`feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards","model testing","competing risks", "prediction"),
           `Cox` = c("semi-parametric","left hand side of the equation", "step function", "interaction of covariates with time"," ", "difficult", "Kaplan-Meier-based"),
           `Case Base Sampling` = c("fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
           "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
           "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities")), escape = FALSE)
```

\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot. Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.
